### 2017-01-10 doc/manuscript.Rmd
Polish up methods

{{ Use an external source via DOI for reference to these methods? }}

Recruitment for the Prospective Metabolism and Islet Cell Evaluation (PROMISE) 
cohort baseline visit took place between 2004-2006 from London and Toronto, 
Canada. Individuals were selected to participant if they met the eligibility 
criteria of having one or more risk factors for type 2 diabetes mellitus, 
including obesity, hypertension, family history of diabetes, and/or a history of
gestational diabetes or birth of a macrosomic infant. A total of 736 individuals
attended the baseline visit. Subsequent examinations occurred every three years,
with three examination visits completed to date (2004-2006, 2007-2009, and
2010-2013). Participants are contacted annually by telephone. The current study
sample used data on participants who did not have diabetes at baseline, who
returned for one or more of the follow-up examinations, and who had samples
available for fatty acid measurements (n=477; see the CONSORT diagram in 
`r cite_sf('consort')`; a total of `r calc_n_for_visits()[1]` attended all three
visits and had complete data). Metabolic characterization, anthropometric
measurements, and questionnaires on lifestyle and sociodemographics are
collected at each examination visit. Research ethics approval was obtained from
Mount Sinai Hospital and the University of Western Ontario, and all participants
provided written informed consent. Data collection methods were standardized
across the 2 centres and research nurses were centrally trained.

@@

Participants completed an 8-12 hour 75g oral glucose tolerance test (OGTT) at 
each examination visit, with blood samples during at fasting, 30 min, and 2 hr 
post-glucose load. Samples were subsequently processed and frozen at -70°C.
Alanine aminotransferase (ALT) was measured using standard laboratory 
procedures. Cholesterol, HDL, and clinically-measured triacylglycerides (TAG) 
were measured using Roche Modular's enzymatic colorimetric tests (Mississauga, 
ON).  Both insulin and glucose were derived from the OGTT at fasting, 30 minute,
and 2 hour time points. Specific insulin was measured with the Elecsys 1010 
(Roche Diagnostics, Basel, Switzerland) immunoassay analyzer and 
electrochemiluminescence immunoassay, which shows 0.05% cross-reactivity to 
intact human pro-insulin and the Des 31,32 circulating split form (Linco Res. 
Inc) and has a coefficient of variation (CV) of 9.3%. Glucose was determined 
using an enzymatic hexokinase (Roche Modular, Roche Diagnostics) with a 
detection range of 0.11 to 41.6 mmol/L and an inter-assay %CV of <1.1% and an
intra-assay %CV of < 1.9%.  All assays were performed at the Banting and Best
Diabetes Centre Core Lab at Mt Sinai Hospital. Impaired fasting glucose (IFG),
impaired glucose tolerance (IGT), and diabetes were categorized using the 2006
WHO criteria [@WHO2006].

TAGFA composition was quantified using stored fasting serum samples from the 
baseline visit, which had been frozen at -70°C for 4-6 years and had not been 
exposed to any freeze-thaw cycles. Serum fatty acids have been documented to be 
stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
of {{ confirm with Richard }} heptadecanoic acid (17:0) was added as an internal
standard prior to extracting total lipids according to the method of Folch 
[@Folch1957a]. Each serum lipid fraction---non-esterified fatty acids (NEFA), 
cholesteryl ester, phospholipid, and TAGFA---was isolated using thin layer 
chromatography. Fatty acid methyl esters were separated and quantified using a 
Varian-430 gas chromatograph (Varian, Lake Forest, CA, USA) equipped with a 
Varian Factor Four capillary column and a flame ionization detector. Fatty acid 
concentrations (nmol/ml) were calculated by proportional comparison of gas
chromatography peak areas to that of the internal standards [@Nishi2014a]. There
were 22 fatty acids measured in the TAGFA fraction. Given that each lipid
fraction differs in biology and given the complexity of the analyses, findings
for other lipid fractions in this cohort are reported separately (see ref
@Johnston2016a and for the analysis of the phospholipid and cholesteryl ester
fractions and {{next citation}} for the analysis of the NEFA fraction).

@@

Covariates were selected based on the previous literature, from directed acyclic
graph [@Greenland1999a] recommendations (see `r cite_st('dag')`), and from 
QIC {{citation?}}. `r cite_st('qic')` shows the covariates compared using QIC.
While the final GEE model selected as best fitting differed between insulin
sensitivity and beta-cell function measures, the differences in QIC values were
less than 10 between the models, suggesting similar fit. As such, we selected
the model that had the fewest covariates and that had similar fit between the
outcome measures ({{##}}). The final GEE model was adjusted for {{covariates}}.
After scaling, log-transforming, and exponentiating, the GEE estimates are 
interpreted as an expected percent difference in the outcome variable for every 
standard deviation (SD) increase in the predictor variable given the covariates 
are held constant. Lastly for the GEE models, we tested for an interaction with 
sex, ethnicity, or time by the predictor term for each outcome variable.

While GEE accounts for the longitudinal design of the data, this approach is 
limited in that it cannot analyze the inherent multivariate nature of the 
composition of the TAGFA fraction. To confirm the GEE results in a multivariate 
environment (all TAGFA), partial least squares regression (PLS) {{citation}} was
used to identify the patterns of TAGFA composition against insulin sensitivity
and beta-cell function. No prevalent diabetes cases were included in the PLS
analysis. For more detailed explanation of GEE and PLS, please see the
{{supplemental}} Methods.

All analyses were performed using R `r getRversion()` [@Rbase], along with the R
packages geepack `r packageVersion('geepack')` for GEE [@Hoejsgaard2006a] and pls
`r packageVersion('pls')` for PLS {{cite?}}. The R code and extra analyses for
this manuscript is available at {{code doi}}. Results were considered
statistically significance at p<0.05, after adjusting for multiple testing using
the Benjamini-Hochberg False Discovery Rate [@Benjamini1995a]. STROBE was used
as a guideline for reporting [@Vandenbroucke2007b].

>>>>

<!-- Use an external source via DOI for reference to these methods? -->

Recruitment for the baseline visit of the Prospective Metabolism and Islet Cell Evaluation (PROMISE) 
cohort took place between 2004-2006 from London and Toronto, 
Canada. Individuals were selected to participant if they met the eligibility 
criteria of having one or more risk factors for type 2 diabetes mellitus, 
including obesity, hypertension, family history of diabetes, and/or a history of
gestational diabetes or birth of a macrosomic infant. A total of 736 individuals
attended the baseline visit. Subsequent examinations occurred every three years,
with three examination visits completed to date (2004-2006, 2007-2009, and
2010-2013). Participants are contacted annually by telephone. The current study
sample used data on participants who did not have diabetes at baseline, who
returned for one or more of the follow-up examinations, and who had samples
available for fatty acid measurements (n=477; see the CONSORT diagram in 
`r cite_sf('consort')`; a total of `r calc_n_for_visits()[1]` attended all three
visits). Metabolic characterization, anthropometric
measurements, and questionnaires on lifestyle and sociodemographics are
collected at each examination visit. Research ethics approval was obtained from
Mount Sinai Hospital and the University of Western Ontario, and all participants
provided written informed consent. Data collection methods were standardized
across the 2 centres and research nurses were centrally trained.

@@

Participants completed an 8-12 hour 75g oral glucose tolerance test (OGTT) at 
each examination visit, with blood samples taken during fasting, 30 min, and 2 hr 
post-glucose load. Samples were subsequently processed and frozen at -70°C.
Alanine aminotransferase (ALT) was measured using standard laboratory 
procedures. Cholesterol, HDL, and clinically-measured triacylglycerides (TAG) 
were measured using Roche Modular's enzymatic colorimetric tests (Mississauga, 
ON).  Both insulin and glucose were derived from the OGTT at fasting, 30 minute,
and 2 hour time points. Specific insulin was measured with the Elecsys 1010 
(Roche Diagnostics, Basel, Switzerland) immunoassay analyzer and 
electrochemiluminescence immunoassay, which shows 0.05% cross-reactivity to 
intact human pro-insulin and the Des 31,32 circulating split form (Linco Res. 
Inc) and has a coefficient of variation (CV) of 9.3%. Glucose was determined 
using an enzymatic hexokinase (Roche Modular, Roche Diagnostics) with a 
detection range of 0.11 to 41.6 mmol/L and an inter-assay %CV of <1.1% and an
intra-assay %CV of < 1.9%.  All assays were performed at the Banting and Best
Diabetes Centre Core Lab at Mt Sinai Hospital. Impaired fasting glucose (IFG),
impaired glucose tolerance (IGT), and diabetes were categorized using the 2006
WHO criteria [@WHO2006].

TAGFA composition was quantified using stored fasting serum samples from the 
baseline visit, which had been frozen at -70°C for 4-6 years and had not been 
exposed to any freeze-thaw cycles. Serum fatty acids have been documented to be 
stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
of {{ confirm with Richard the correct name}} heptadecanoic acid (17:0) was added as an internal
standard prior to extracting total lipids according to the method of Folch 
[@Folch1957a]. Each serum lipid fraction---non-esterified fatty acids (NEFA), 
cholesteryl ester, phospholipid, and TAGFA---was isolated using thin layer 
chromatography. Fatty acid methyl esters were separated and quantified using a 
Varian-430 gas chromatograph (Varian, Lake Forest, CA, USA) equipped with a 
Varian Factor Four capillary column and a flame ionization detector. Fatty acid 
concentrations (nmol/ml) were calculated by proportional comparison of gas
chromatography peak areas to that of the internal standards [@Nishi2014a]. There
were 22 fatty acids measured in the TAGFA fraction. Given that each lipid
fraction differs in biology and given the complexity of the analyses, findings
for other lipid fractions in this cohort are reported separately (see ref
@Johnston2016a for the analysis of the phospholipid and cholesteryl ester
fractions). 
<!--and {{next citation}} for the analysis of the NEFA fraction).-->

@@

Covariates were selected based on the previous literature, from directed acyclic
graph [@Greenland1999a] recommendations (see `r cite_st('dag')`), and from 
QIC. `r cite_st('qic')` shows the covariates compared using QIC.
While the final GEE model selected as best fitting differed between insulin
sensitivity and beta-cell function measures, the differences in QIC values were
less than 10 between many of the models, suggesting similar fit. As such, we selected
the model that had the fewest covariates and that had similar fit between the
outcome measures ({{##}}). The final GEE model was adjusted for {{covariates}}.
After scaling, log-transforming, and exponentiating, the GEE estimates are 
interpreted as an expected percent difference in the outcome variable for every 
standard deviation (SD) increase in the predictor variable given the covariates 
are held constant (including time). Lastly for the GEE models, we tested for an interaction with 
sex, ethnicity, or time by the predictor term for each outcome variable.

While GEE accounts for the longitudinal design of the data, this approach is 
limited in that it cannot analyze the inherent multivariate nature of the 
composition of the TAGFA fraction. To confirm the GEE results in a multivariate 
environment (all TAGFA), partial least squares regression (PLS) was
used to identify the patterns of TAGFA composition against insulin sensitivity
and beta-cell function. No prevalent diabetes cases were included in the PLS
analysis. For more detailed explanation of GEE and PLS, please see the
{{supplemental}} Methods.

All analyses were performed using R `r getRversion()` [@Rbase], along with the R
packages geepack `r packageVersion('geepack')` for GEE [@Hoejsgaard2006a] and pls
`r packageVersion('pls')` for PLS. The R code and extra analyses for
this manuscript is available at {{code doi}}. Results were considered
statistically significance at p<0.05, after adjusting for multiple testing using
the Benjamini-Hochberg False Discovery Rate [@Benjamini1995a]. STROBE was used
as a guideline for reporting [@Vandenbroucke2007b].
